A thermodynamic view of dusty protoplanetary disks

### 2017-08-02 linear.tex
revising in response to AY comments, up to section 5.1.1

\section{Application to protoplanetary disks}\label{linear_problem}
We now examine instabilities in dusty protoplanetary disks
based on explicit descriptions of the equilibrium state. We first show
that sharp radial edges in the dust component are unstable. We then
revisit two well-known instabilities in dusty gas using the
hydrodynamic framework developed thusfar, namly secular graviational
instabilities and the streaming instability.

@@

We usually consider gas-dominated disks with $\epsilon_0 < 1$.  
Then $\rhog(r,z)$ is effectively Gaussian, as in the 
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \epsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z^2}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

@@

\subsubsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

>>>>

\section{Application to protoplanetary disks}\label{linear_problem}
We now examine instabilities in dusty protoplanetary disks
based on explicit descriptions of the equilibrium state. We first
specify explicit disk structures to analyze. We show 
that sharp radial edges in the dust component are unstable. We then
revisit two well-known instabilities in dusty gas using the
hydrodynamic framework developed thusfar, namly secular graviational
instabilities and the streaming instability.

@@

In gas-dominated disks with $\epsilon_0 < 1$ the gas distribution
$\rhog(r,z)$ is close to Gaussian, as in the   
dust-free case, and the dust density is approximately 
\begin{align}
  \rhod \simeq \epsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z^2}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height.% In numerical
%calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
%$\Htilde$ for input.

@@

\subsubsection{Orbital frequency} 
From Eq. \ref{steady_momr}} the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

### 2017-08-02 model.tex
revising in response to AY comments, up to section 5.1.1

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}.% obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

@@

In this section we discuss the stability properties of the 
dusty-gas mixture using a variational principle. This method does not
require explicit solutions to the equilibrium equations. Self-gravity is neglected here.

\subsection{Steady states}\label{eqm}
We consider axisymmetric steady states as background equilibria to
perturb. For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in 
\S\ref{steady_state}, when we explicitly analyse the stability of
protoplanetary disks.

The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Since $\seff$ depends on $\tepsilon$, we see that radial gradients in
the dust-to-gas ratio also drives vertical shear \citep[cf. that 
driven by $\p_z\epsilon$ usually discussed in dusty disks,  
e.g.][]{chiang08}.

However, disk structures obtained from
Eq. \ref{steady_momr}---\ref{steady_momz} generally do not satisfy  
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instabilities of interest.

@@

%Our discussion is based on
%the corresponding axisymmetric linearized equations. 
We consider axisymmetric Eulerian perturbations to a variable $X$ of
the form  
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Then perturbations have time dependence $e^{st +
  \ii\omega t}$.  Thus for $\omega>0$, perturbations rotate
anti-clockwise in the complex plane.

In Appendix \ref{var_prin} we derive the integral relation from the
corresponding linearized equations (Eqs.\ref{masseq}--\ref{momeq}, and 
Eq. \ref{eff_energy} in place of the dust continuity equation): 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

\subsection{Strictly isothermal gas perfectly coupled to dust}\label{iso_perfect}
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}. 
We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas.

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\tepsilon =
0$) but  radially-varying ($\p_r\tepsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-c_s^4\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} but there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$)
that can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-to-gas ratio. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

@@

Consider oscillations in the dusty gas with period $T_p$. 
The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{D\upsilon}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep{cox67}, 
where $\upsilon=1/\rho$ is the specific volume of the mixture, $D/Dt$
is the Lagragian derivative, and the 
spatial integral is taken over the total mass $M$ of the mixture and
$dm$ is a mass element.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
%  s = \frac{\imag\int P
%  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd 
%    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check} 
  s = \frac{\imag\int P
    \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\mathcal{I}^2}, \label{vsi_check} 
\end{align} 
again assuming $\omega\neq0$. 
This is VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We present 
numerical solutions of the VSI with perfectly-coupled dust in \S\ref{results}.

>>>>

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

@@

In this section we discuss the stability properties of the 
dusty-gas mixture using a variational principle. This approach does not
require explicit solutions (i.e. a specific distribution of the density and dust-to-gas ratio)  
to the equilibrium equations. We consider axisymmetric systems and neglect self-gravity here.

\subsection{Steady states}\label{eqm}
%We consider axisymmetric steady states as background equilibria to
%perturb.

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
solutions with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in 
\S\ref{steady_state}, when we analyse the stability of
protoplanetary disks.

The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   &= \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right).% \\
%& = \frac{1}{\rhog}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
%    P}{\p z}\frac{\p \seff}{\p r} \right)
\end{align}
and recall $\seff = \ln{\left[c_s^2\left(1-\tepsilon\right)\right]}$. It is well appreciated that 
vertical stratification of the dust layer contributes to vertical shear
\citep{chiang08}. Eq. \ref{vshear} shows that a radial dust
stratification ($\p_r\tepsilon$) also contributes to vertical shear.

%Since $\seff$ depends on $\tepsilon$, we see that radial gradients in
%the dust-to-gas ratio also drives vertical shear \citep[cf. that 
%driven by $\p_z\epsilon$ usually discussed in dusty disks,  
%e.g.][]{chiang08}.

Our equilibrium solutions satisfy the hydrostatic constraints of
Eq. \ref{steady_momr}---\ref{steady_momz} but are not in general
steady state solutions to the energy 
equation (\ref{eff_energy}), because the source term $\mathcal{C} \ne0$ for realistic disks.  
Limiting cases with $\mathcal{C} \equiv 0$ include \begin{inparaenum}[1)] 
\item 
  unstratified or 2D, razor-thin disk models with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
The following analyses thus apply strictly to these limiting cases of
$\mathcal{C} = 0$.  Our analysis is also a good approximation when
$\tstop$ is sufficiently small such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instabilities of interest.

%However, disk structures obtained from
%Eq. \ref{steady_momr}---\ref{steady_momz} generally do not satisfy  
%the steady state effective energy Eq. \ref{eff_energy} because typically 
%$\mathcal{C}\neq0$.  In order to have a well-defined stability problem, we formally
%consider equilibria with $\mathcal{C}\equiv 0$. Namely      
%\begin{inparaenum}[1)] 
%\item 
%  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
%\item 
%  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
%\end{inparaenum} 
%Alternatively, we are assuming that $\tstop$  is sufficiently small

@@

%Our discussion is based on
%the corresponding axisymmetric linearized equations. 
We consider axisymmetric Eulerian perturbations to a variable $X$ of
the form  
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. We 
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Then perturbations have time dependence $e^{st +
  \ii\omega t}$.  Thus for $\omega>0$, perturbations rotate
anti-clockwise in the complex plane.

In Appendix \ref{var_prin} we derive linearize Eqs. \ref{masseq}, \ref{momeq}, and 
Eq. \ref{eff_energy} to derive the following integral relation: 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

\subsection{Strictly isothermal gas perfectly coupled to dust}\label{iso_perfect}
When $c_s^2$ is a constant and $\tstop=0$ (so that $\mathcal{C} = 0$), the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit
adiabatic 
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + \frac{1}{\rhog}\nabla P \cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -\frac{1}{\rhog}\frac{\p P}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
%  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
%    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
%\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. Note that $\nabla P/\rhog =
c_s^2\nabla \ln{\rhog}$ for strictly isothermal gas considered here.

Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_P\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}. This substitution is
consistent with our definition of the effective entropy $\seff$ since
we are considering constant $c_s$.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas.

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed but radially-stratified,  
such that $\p_z\tepsilon = 0$ but $\p_r\tepsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\left(\frac{1}{\rhog}\frac{\p P}{\p
    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
\end{align}
%\begin{align}
%-c_s^4\left(\frac{\p\ln{\rhog}}{\p
%    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
%\end{align}
Such an isothermal dusty disk is \emph{unstable} because there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$),
which can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-fraction. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
The physical interpretation of \S\ref{grow_osc} is here derived in
detail. Consider constant $c_s$ but $\tstop\neq0$ (so $\delta\mathcal{C}
\neq 0$). Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
represent correlations between compression/expansion and  
heating/cooling.

@@

Consider oscillations in the dusty gas with period $T_p$. 
The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{D\upsilon}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep[][Eq. 4.10]{cox67}, 
where $\upsilon=1/\rho$ is the specific volume of the mixture, $D/Dt$
is the Lagragian derivative, and the 
spatial integral is taken over the total mass $M$ of the mixture and
$dm$ is a mass element.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
%  s = \frac{\imag\int P
%  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd 
%    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check} 
  s = \frac{\imag\int P
    \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\mathcal{I}^2}, \label{vsi_check} 
\end{align} 
again assuming $\omega\neq0$. 
This instability represents VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We present 
numerical solutions of the VSI with perfectly-coupled dust in \S\ref{results}.
